---
title: "Ruthenibacterium functional comparison"
author: "Avril Metcalfe-Roach"
date: "2026-03-10"
output: html_document
---

Pulling all the ruthenibacterium functions to see what differentiates it from other taxa.
CURRENTLY REMOVING THE UNINTEGRATED COUNTS EVEN IF IT'S ANNOTATED AS RUTHENIBACTERIUM

```{r}
library(tidyverse)
library(phyloseq)
path = data.table::fread('../Data/humann_exports_mgx_g4h/mgx_genefamilies_stratified.tsv', check.names = F)
path_sep = path %>% 
  separate(`# Gene Family`, into = c('feature', 'taxon'), sep = '\\|')
  
# Select ruthenibacterium
ruth = path_sep %>% filter(str_detect(taxon, 'Ruthenibacterium'),
                           feature !='UNINTEGRATED') %>% 
  select(-taxon) %>% 
  column_to_rownames(var = 'feature')

# Let's remove all columns that have only 0 values (these don't have the taxon present)
w = which(colSums(ruth) ==0)
ruth = ruth[,-w]

# Rarefy this count data table
ruth_rare = phyloseq(otu_table(round(ruth,0), taxa_are_rows = TRUE)) %>% rarefy_even_depth(rngseed = 421)
```

We will create a PCoA plot to differentiate samples based on functional similarity.

```{r}
# First, we need to calculate the distance matrix. We'll use Bray-Curtis distance for this.
library(vegan)

make_plot = function(df,method){
  # df=ruth_rare; method='bray'
  
  ruth_dist = vegdist(t(df), method = method)

  # Now we can perform PCoA on the distance matrix
  ruth_pcoa = cmdscale(ruth_dist, k = 2, eig = TRUE)
  # Let's create a data frame for plotting
  pcoa_df = data.frame(Sample = rownames(ruth_pcoa$points),
                        PC1 = ruth_pcoa$points[,1],
                        PC2 = ruth_pcoa$points[,2])
  
  
  # Correlate functions with the axes
  loadings_pc1 <- apply(data.frame(df), 1, function(feature_abund) {
    cor(pcoa_df$PC1, feature_abund) # Just take first sample vector match
  })
  loadings_pc2 <- apply(data.frame(df), 1, function(feature_abund) {
    cor(pcoa_df$PC2, feature_abund) # Just take first sample vector match
  })
  loadings = tibble(feature = rownames(df), 
                    PC1_loading = loadings_pc1, 
                    PC2_loading = loadings_pc2) %>%
    arrange(desc(abs(PC1_loading)))
  
  # Now we can plot the PCoA
  p = ggplot(pcoa_df, aes(x = PC1, y = PC2)) +
    geom_point(size = 3) +
    labs(title = paste("R. lactatiformans functions - ", method),
         x = paste0("PC1 (", round(ruth_pcoa$eig[1] / sum(ruth_pcoa$eig) * 100, 2), "%)"),
         y = paste0("PC2 (", round(ruth_pcoa$eig[2] / sum(ruth_pcoa$eig) * 100, 2), "%)")) +
    theme_minimal(base_size=18)
  
  print(p)
  return(list(plot=p,loadings=loadings))
}

p_bray = make_plot(ruth_rare, method = 'bray')
p_jaccard = make_plot(ruth_rare, method = 'jaccard')
```

Now take a look at the top loadings

```{r}
p_bray_pc1 = p_bray$loadings %>% slice_head(n=20) # richness and evenness
p_jaccard_pc1 = p_jaccard$loadings %>% slice_head(n=20) # richness

sig_ids = intersect(p_bray_pc1$feature, p_jaccard_pc1$feature)  # 19 shared
sig_ids

# Manual annotation because the website isn't working
# Lots of these are associated with Subdoligranulum, another Oscillospiraceae
sig_ids_annot = tibble(feature= sig_ids,
                       annotation = c("Cluster: ABC transporter, ATP-binding protein",
                                      "Cluster: Multidrug-efflux transporter",
                                      "Cluster: Uncharacterized protein",
                                      "tRNA-t(6)A37 methylthiotransferase",
                                      "Cluster: Stage 0 sporulation protein A homolog",
                                      "Cluster: Radical SAM protein",
                                      "Cluster: Uncharacterized protein",
                                      "RNA polymerase sigma factor, region 3/4-like",
                                      "HAMP domain-containing protein",
                                      "Uncharacterized protein",
                                      "Cluster: GHMP kinase, N-terminal domain protein",
                                      "MATE family efflux transporter",
                                      "Sugar (Glycoside-Pentoside-Hexuronide) transporter",
                                      "N-acetyltransferase domain-containing protein",
                                      "Cluster: Membrane transport protein MMPL domain-containing protein",
                                      "Uncharacterized protein",
                                      "Cluster: Nicotinate phosphoribosyltransferase",
                                      "Cluster: L-serine deaminase",
                                      "Flavodoxin")) %>% 
  left_join(p_bray_pc1 %>% select(-PC2_loading))
sig_ids_annot[sig_ids_annot$PC1_loading > 0,] %>% arrange(desc(PC1_loading))
sig_ids_annot[sig_ids_annot$PC1_loading < 0,] %>% arrange(desc(PC1_loading))
```

# Explore using the x axis of the PCA plot to categorize the samples

```{r}
# Load phyloseq objct
ps = readRDS('../Data/pilot_metagenomic_phyloseq_g4h.rds')
sam = data.frame(ps@sam_data) %>% 
  rownames_to_column('Sample')

# Fix names, add metadata ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# sample key
key <- read_csv("~/Library/CloudStorage/OneDrive-UBC/CMMI/ANALYSES/Pilot/Data/sample_key.csv")

pcoa_df = p_bray$plot$data %>% `rownames<-`(NULL) %>% 
  select(Sample,PC1) %>% 
  mutate(old = sapply(.$Sample,function(x)str_split(x, pattern = "_",)[[1]][1] %>% as.numeric())) %>% 
  left_join(key) %>%
  select(-old, -Sample) %>%
  rename(Sample=new) %>% 
  left_join(sam)
```
 NUMERIC
```{r}
# Identify all the numeric columns
numeric_cols <- sapply(pcoa_df, is.numeric)

# Create a new data frame with only the numeric columns
pcoa_numeric <- pcoa_df[, numeric_cols]

# Correlate the numeric columns with PC1
cor_results <- sapply(pcoa_numeric, function(x) cor(x, pcoa_df$PC1, method = "spearman"))

cor_results_df <- data.frame(Variable = names(cor_results), Correlation = cor_results) %>%
  arrange(desc(abs(Correlation)))
```

```{r}
# Same plot as above, but for all variables with a correlation over 0.3
to_plot = cor_results_df %>% filter(abs(Correlation) > 0.3) %>% pull(Variable)
for (var in to_plot) {
  p <- ggplot(pcoa_df, aes_string(x = var, y = "PC1")) +
    geom_point() +
    geom_smooth(method = 'lm') +
    labs(title = paste("PC1 vs", var),
         y = "PC1 Score",
         x = var) +
    theme_minimal(base_size=18)
  
  print(p)
}
```

Let's use teh same approach, but for the categorical variables.
```{r}
# Identify all the categorical columns
categorical_cols <- sapply(pcoa_df, is.character)
# Create a new data frame with only the categorical columns
pcoa_categorical <- pcoa_df[, categorical_cols]
# Create boxplots for each categorical variable against PC1
for (var in names(pcoa_categorical)) {
  p <- ggplot(pcoa_df, aes_string(x = var, y = "PC1")) +
    geom_boxplot() +
    labs(title = paste("PC1 vs", var),
         y = "PC1 Score",
         x = var) +
    theme_minimal(base_size=18)
  
  print(p)
}
```


