---
title: "Ruthenibacterium functional comparison"
author: "Avril Metcalfe-Roach"
date: "2026-03-10"
output: html_document
---

Pulling all the ruthenibacterium functions to see what differentiates it from other taxa.
CURRENTLY REMOVING THE UNINTEGRATED COUNTS EVEN IF IT'S ANNOTATED AS RUTHENIBACTERIUM

```{r}
library(tidyverse)
path = data.table::fread('../Data/humann_exports_mgx_g4h/mgx_genefamilies_stratified.tsv', check.names = F)
path_sep = path %>% 
  separate(`# Gene Family`, into = c('feature', 'taxon'), sep = '\\|')
  

ruth = path_sep %>% filter(str_detect(taxon, 'Ruthenibacterium'),
                           feature !='UNINTEGRATED')
unique(ruth$taxon) # only ruthenibacterium lactatiformans present, good
```
Prevalence of Ruthenibacterium lactatiformans in cohort is 68%.

```{r}
ruth_prev = table(Present=colSums(ruth[,3:ncol(ruth)]) > 0) / (ncol(ruth)-2)
ruth_prev %>% round(2) # 68% prevalent
```

## Abundance-related analysis
Determining the functional spreads of Ruthenibacterium lactatiformans across the cohort.
```{r}
# First, we'll transform the counts into relative abundance per sample
ruth_rel = ruth
ruth_rel[,3:ncol(ruth_rel)] = apply(ruth_rel[3:ncol(ruth_rel)], 2, function(x) x / sum(x, na.rm = TRUE))

# Let's remove all columns that have only NAN values (these don't have the taxon present)
w = which(colSums(is.na(ruth_rel)) == nrow(ruth_rel))
ruth_rel[,w] %>% head() # works as intended
ruth_rel = ruth_rel[,-w]

# Now we can calculate the mean relative abundance of each function across the samples where Ruthenibacterium lactatiformans is present
ruth_rel_pivot = ruth_rel %>%
  pivot_longer(cols = 3:ncol(ruth_rel), names_to = "Sample", values_to = "RelativeAbundance") %>%
  group_by(feature) %>%
  summarise(MeanRelativeAbundance = mean(RelativeAbundance, na.rm = TRUE),
            SDRelativeAbundance = sd(RelativeAbundance, na.rm = TRUE)) %>%
  mutate(CVRelativeAbundance = abs(SDRelativeAbundance / MeanRelativeAbundance)) %>%
  arrange(desc(CVRelativeAbundance))
head(ruth_rel_pivot) # A few pathways that are somewhat variable across the samples

# Too many hits. 
# Let's take things that are:
#   1. at least 1/1000 of the total ruthenibacterium reads.
#   2. a CV of at least 2.
ruth_filt = ruth_rel_pivot %>% 
  filter(MeanRelativeAbundance > 0.001) %>%
  filter(CVRelativeAbundance > 2) %>%
  arrange(desc(CVRelativeAbundance))

# Coefficient of variation bar plot
ggplot(ruth_filt, 
       aes(x = reorder(feature, CVRelativeAbundance), y = CVRelativeAbundance)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "UniRef50, >0.1% rel abundance, CV > 2",
       x = "Function",
       y = "Coefficient of Variation") +
  theme_minimal()

# Now try the same thing, but boxplots
ruth_rel %>%
  filter(feature %in% ruth_filt$feature) %>%
  pivot_longer(cols = 3:ncol(ruth_rel), names_to = "Sample", values_to = "RelativeAbundance") %>%
  mutate(feature = factor(feature, levels = rev(unique(ruth_filt$feature)))) %>%
  ggplot(aes(x = feature, y = RelativeAbundance)) +
  geom_boxplot(fill='lightgreen') +
  coord_flip() +
  labs(title = "UniRef50, >0.1% rel abundance, CV > 2",
       x = "Function",
       y = "Relative Abundance") +
  theme_minimal()
```

## Prevalence-related analysis

```{r}
prev = ruth_rel %>%
  pivot_longer(cols = 3:ncol(ruth_rel), names_to = "Sample", values_to = "RelativeAbundance") %>% 
  mutate(Present = RelativeAbundance > 0) %>%
  group_by(feature) %>%
  summarise(Prevalence = mean(Present, na.rm = TRUE)) %>%
  ungroup() #%>% 
  # filter(feature %in% ruth_filt$feature) %>%
  # mutate(feature = factor(feature, levels = rev(unique(ruth_filt$feature))))

ggplot(prev, aes(x = feature, y = Prevalence)) +
  geom_bar(stat = "identity", fill = "red3") +
  coord_flip() +
  labs(x = "Function",
       y = "Prevalence") +
  theme_minimal()

ggplot(prev %>% 
         mutate(feature = factor(feature, levels = rev(unique(ruth_rel_pivot_present$feature)))), 
       aes(x = feature, y = Prevalence)) +
  geom_bar(stat = "identity", fill = "red3") +
  coord_flip() +
  labs(x = "Function",
       y = "Prevalence") +
  theme_minimal()
```

# Let's look at abundance again, but we will only average across the samples where the function is present
# to get a better idea of the abundance when it's actually there

```{r}
ruth_rel_pivot_present = ruth_rel %>%
  pivot_longer(cols = 3:ncol(ruth_rel), names_to = "Sample", values_to = "RelativeAbundance") %>%
  # filter(feature %in% ruth_filt$feature) %>%
  group_by(feature) %>%
  summarise(MeanRelativeAbundance = mean(RelativeAbundance[RelativeAbundance > 0], na.rm = TRUE),
            SDRelativeAbundance = sd(RelativeAbundance[RelativeAbundance > 0], na.rm = TRUE)) %>%
  mutate(CVRelativeAbundance = abs(SDRelativeAbundance / MeanRelativeAbundance)) %>%
  arrange(desc(CVRelativeAbundance))

ggplot(ruth_rel_pivot_present, 
       aes(x = reorder(feature, CVRelativeAbundance), y = CVRelativeAbundance)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(x = "Function",
       y = "Coefficient of Variation (when present)") +
  theme_minimal()

# Boxplot of abundances
ruth_rel %>%
  pivot_longer(cols = 3:ncol(ruth_rel), names_to = "Sample", values_to = "RelativeAbundance") %>%
  filter(feature %in% ruth_filt$feature) %>%
  filter(RelativeAbundance > 0) %>%
  mutate(feature = factor(feature, levels = rev(unique(ruth_rel_pivot_present$feature)))) %>%
  ggplot(aes(x = feature, y = RelativeAbundance)) +
  geom_boxplot(fill='lightgreen') +
  coord_flip() +
  labs(x = "Function",
       y = "Relative Abundance (when present)") +
  theme_minimal()

# Now let's look at the relationship between prevalence and abundance for each function
prev_abund = prev %>%
  left_join(ruth_rel_pivot_present, by = "feature")
ggplot(prev_abund, aes(y = Prevalence, x = MeanRelativeAbundance)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(y = "Prevalence",
       x = "Mean Relative Abundance (when present)") +
  theme_minimal()

# Coefficient of variation vs mean relative abundance
ggplot(prev_abund, aes(x = MeanRelativeAbundance, y = CVRelativeAbundance)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "Mean Relative Abundance (when present)",
       y = "Coefficient of Variation (when present)") +
  theme_minimal()
```
