---
title: "Ruthenibacterium functional comparison"
author: "Avril Metcalfe-Roach"
date: "2026-03-10"
output: html_document
---

Pulling all the ruthenibacterium functions to see what differentiates it from other taxa.
CURRENTLY REMOVING THE UNINTEGRATED COUNTS EVEN IF IT'S ANNOTATED AS RUTHENIBACTERIUM

```{r}
library(tidyverse)

path = read.delim('Data/humann_exports_mgx_g4h/mgx_pathabundance_stratified.tsv', check.names = F)
path_sep = path %>% 
  separate(`# Pathway`, into = c('feature', 'taxon'), sep = '\\|')
  

ruth = path_sep %>% filter(str_detect(taxon, 'Ruthenibacterium'),
                           feature !='UNINTEGRATED')
unique(ruth$taxon) # only ruthenibacterium lactatiformans present, good
```
Prevalence of Ruthenibacterium lactatiformans in cohort is 68%.

```{r}
ruth_prev = table(Present=colSums(ruth[,3:ncol(ruth)]) > 0) / (ncol(ruth)-2)
ruth_prev %>% round(2) # 68% prevalent
```

## Abundance-related analysis
Determining the functional spreads of Ruthenibacterium lactatiformans across the cohort.
```{r}
# First, we'll transform the counts into relative abundance per sample
ruth_rel = ruth
ruth_rel[,3:ncol(ruth_rel)] = apply(ruth_rel[3:ncol(ruth_rel)], 2, function(x) x / sum(x, na.rm = TRUE))

# Let's remove all columns that have only NAN values (these don't have the taxon present)
w = which(colSums(is.na(ruth_rel)) == nrow(ruth_rel))
ruth_rel[,w] %>% head() # works as intended
ruth_rel = ruth_rel[,-w]

# Now we can calculate the mean relative abundance of each function across the samples where Ruthenibacterium lactatiformans is present
ruth_rel_pivot = ruth_rel %>%
  pivot_longer(cols = 3:ncol(ruth_rel), names_to = "Sample", values_to = "RelativeAbundance") %>%
  group_by(feature) %>%
  summarise(MeanRelativeAbundance = mean(RelativeAbundance, na.rm = TRUE),
            SDRelativeAbundance = sd(RelativeAbundance, na.rm = TRUE)) %>%
  mutate(CVRelativeAbundance = abs(SDRelativeAbundance / MeanRelativeAbundance)) %>%
  arrange(desc(CVRelativeAbundance))
head(ruth_rel_pivot) # A few pathways that are somewhat variable across the samples

# Coefficient of variation bar plot
ggplot(ruth_rel_pivot, 
       aes(x = reorder(feature, CVRelativeAbundance), y = CVRelativeAbundance)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(x = "Function",
       y = "Coefficient of Variation") +
  theme_minimal()

# Now try the same thing, but boxplots
ruth_rel %>%
  pivot_longer(cols = 3:ncol(ruth_rel), names_to = "Sample", values_to = "RelativeAbundance") %>%
  mutate(feature = factor(feature, levels = rev(unique(ruth_rel_pivot$feature)))) %>%
  ggplot(aes(x = feature, y = RelativeAbundance)) +
  geom_boxplot(fill='lightgreen') +
  coord_flip() +
  labs(x = "Function",
       y = "Relative Abundance") +
  theme_minimal()


# Functional bar plot, cluster the samples and functions according to their similarity
library(vegan)
temp = ruth_rel[,3:ncol(ruth_rel)] %>% `rownames<-`(ruth_rel$feature)

# SAMPLE CLUSTERING
dist_matrix <- vegdist(t(temp), method = "bray")
hc <- hclust(dist_matrix, method = "average")
# FUNCTIONAL CLUSTERING
dist_matrix_func <- vegdist(temp, method = "bray")
hc_func <- hclust(dist_matrix_func, method = "average")

# Reorder data based on the clustering
data_reordered <- temp[hc_func$order, hc$order] %>% 
  data.frame() %>% 
  rownames_to_column(var = "feature") %>%
  pivot_longer(cols = 2:ncol(.), names_to = "Sample", values_to = "RelativeAbundance") %>% 
  mutate(feature = factor(feature, levels = unique(feature)),
         Sample = factor(Sample, levels = unique(Sample)))


# Barplot clustered by similarity
ggplot(data_reordered, aes(x = Sample, y = RelativeAbundance, fill = feature)) +
  geom_bar(stat = "identity") +
  labs(x = "Sample",
       y = "Relative Abundance",
       fill = "Function") +
  theme_minimal() +
  theme(legend.position = "none") + # Hide legend for better visualization
  theme(axis.text.x = element_blank())
```

## Prevalence-related analysis

```{r}
prev = data_reordered %>% 
  mutate(Present = RelativeAbundance > 0) %>%
  group_by(feature) %>%
  summarise(Prevalence = mean(Present, na.rm = TRUE)) %>%
  mutate(feature = factor(feature, levels = rev(unique(ruth_rel_pivot$feature))))

ggplot(prev, aes(x = feature, y = Prevalence)) +
  geom_bar(stat = "identity", fill = "red3") +
  coord_flip() +
  labs(x = "Function",
       y = "Prevalence") +
  theme_minimal()

ggplot(prev %>% mutate(feature = factor(feature, levels = rev(unique(ruth_rel_pivot_present$feature)))), 
       aes(x = feature, y = Prevalence)) +
  geom_bar(stat = "identity", fill = "red3") +
  coord_flip() +
  labs(x = "Function",
       y = "Prevalence") +
  theme_minimal()
```

# Let's look at abundance again, but we will only average across the samples where the function is present
# to get a better idea of the abundance when it's actually there

```{r}
ruth_rel_pivot_present = ruth_rel %>%
  pivot_longer(cols = 3:ncol(ruth_rel), names_to = "Sample", values_to = "RelativeAbundance") %>%
  group_by(feature) %>%
  summarise(MeanRelativeAbundance = mean(RelativeAbundance[RelativeAbundance > 0], na.rm = TRUE),
            SDRelativeAbundance = sd(RelativeAbundance[RelativeAbundance > 0], na.rm = TRUE)) %>%
  mutate(CVRelativeAbundance = abs(SDRelativeAbundance / MeanRelativeAbundance)) %>%
  arrange(desc(CVRelativeAbundance))

ggplot(ruth_rel_pivot_present, 
       aes(x = reorder(feature, CVRelativeAbundance), y = CVRelativeAbundance)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(x = "Function",
       y = "Coefficient of Variation (when present)") +
  theme_minimal()

# Boxplot of abundances
ruth_rel %>%
  pivot_longer(cols = 3:ncol(ruth_rel), names_to = "Sample", values_to = "RelativeAbundance") %>%
  filter(RelativeAbundance > 0) %>%
  mutate(feature = factor(feature, levels = rev(unique(ruth_rel_pivot_present$feature)))) %>%
  ggplot(aes(x = feature, y = RelativeAbundance)) +
  geom_boxplot(fill='lightgreen') +
  coord_flip() +
  labs(x = "Function",
       y = "Relative Abundance (when present)") +
  theme_minimal()

# Now let's look at the relationship between prevalence and abundance for each function
prev_abund = prev %>%
  left_join(ruth_rel_pivot_present, by = "feature")
ggplot(prev_abund, aes(y = Prevalence, x = MeanRelativeAbundance)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(y = "Prevalence",
       x = "Mean Relative Abundance (when present)") +
  theme_minimal()

# Coefficient of variation vs mean relative abundance
ggplot(prev_abund, aes(x = MeanRelativeAbundance, y = CVRelativeAbundance)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "Mean Relative Abundance (when present)",
       y = "Coefficient of Variation (when present)") +
  theme_minimal()
```

# Are key pathways present in other taxa?

Testing the "DTDPRHAMSYN-PWY: dTDP-&beta;-L-rhamnose biosynthesis" pathway, which is one of the most abundant YET low prevalence

```{r}
# Convert to relative abundance
path_rel = path_sep %>% filter(feature != 'UNINTEGRATED')
path_rel[,3:ncol(path_rel)] = apply(path_rel[3:ncol(path_rel)], 2, function(x) x / sum(x, na.rm = TRUE))

# Filter for pathway of interest
key_pathway = path_rel %>% 
  filter(feature == "DTDPRHAMSYN-PWY: dTDP-&beta;-L-rhamnose biosynthesis")

# Prevalence of pathway
key_pathway_prev = table(Present=colSums(key_pathway[,3:ncol(key_pathway)]) > 0) / (ncol(key_pathway)-2)
key_pathway_prev # Everyone has it

# boxplot of abundance
boxplot(colSums(key_pathway[,3:ncol(key_pathway)]), 
        main = "Relative Abundance of DTDPRHAMSYN-PWY", 
        ylab = "Relative Abundance", col = "lightblue")
ggplot(aes(x = 1, y = colSums(key_pathway[,3:ncol(key_pathway)]))) +
  geom_boxplot(fill='lightblue') +
  labs(x = "Sample",
       y = "Relative Abundance of DTDPRHAMSYN-PWY") +
  theme_minimal() +
  theme(axis.text.x = element_blank())

# Check which taxa contribute the most reads per sample
most_reads = path_sep %>% 
  filter(feature == "DTDPRHAMSYN-PWY: dTDP-&beta;-L-rhamnose biosynthesis",
         taxon != 'unclassified') %>%
  pivot_longer(cols = 3:ncol(.), names_to = "Sample", values_to = "Abundance") %>%
  group_by(Sample) %>%
  mutate(RelativeAbundance = Abundance / sum(Abundance, na.rm = TRUE)) %>% 
  ungroup() %>% 
  arrange(desc(RelativeAbundance)) %>%
  group_by(Sample) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  select(Sample, taxon, RelativeAbundance) %>%
  filter(RelativeAbundance > 0)

most_reads %>% head() # 
```

